---
title: 'chapter 14 Joint Variability'
output:
  tufte::tufte_html:
    tufte_features: ['fonts','background']
    toc: true
    toc_depth: 2
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = FALSE)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(extraDistr)
library(gridExtra)
library(latex2exp)
library(moments)
library(bookdown)
library(rsconnect)
```

```{r, include=FALSE}
paygap <- read.csv('./data/gender-paygap-2019.csv')
paygap <- paygap %>%
  mutate(EmployerSize = factor(EmployerSize, levels = c('0-249','250-499','500-999','1000-4999','5000-19999','20000+')))
nycheat <- read.csv('./data/nyc-heatwave.csv')
```

\newcommand{\E}{\text{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\SD}{\text{SD}}
\newcommand{\SE}{\text{SE}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\Cor}{\text{Cor}}
\renewcommand{\P}{\text{P}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\sumin}{\sum_i^n}
\newcommand{\Bias}{\text{Bias}}

---

# Covariance 

Covariance quantifies the joint variability of two variables.  

derivation, formula for covariance 





---

# Correlation 

Correlation is a derived quantity. It normalized covariance.  





---

# Multivariate Normal 

We introduced the normal distribution in chapter 9, as the continuous probability distribution with the density function:


A single distribution of data can be modeled using the univariate normal distribution, as given in the equation above.  

When we consider the joint variability of two variables, we can think of them as giving rise to a multivariate normal, i.e. a 3d gaussian:



The multivariate Gaussian has mean vector mu and correlation matrix Sigma.   












