---
title: 'chapter 21 Bias Variance Tradeoff'
output:
  tufte::tufte_html:
    tufte_features: ['fonts','background','italics']
    css: '../mytufte.css'
    toc: true
    toc_depth: 2
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
link-citations: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = FALSE)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(extraDistr)
library(gridExtra)
library(latex2exp)
library(moments)
library(bookdown)
library(rsconnect)
```

```{r, include=FALSE}
paygap <- read.csv('./data/gender-paygap-2019.csv')
paygap <- paygap %>%
  mutate(EmployerSize = factor(EmployerSize, levels = c('0-249','250-499','500-999','1000-4999','5000-19999','20000+')))
nycheat <- read.csv('./data/nyc-heatwave.csv')
```

\newcommand{\E}{\text{E}}
\newcommand{\Var}{\text{Var}}
\newcommand{\SD}{\text{SD}}
\newcommand{\SE}{\text{SE}}
\newcommand{\Cov}{\text{Cov}}
\newcommand{\Cor}{\text{Cor}}
\renewcommand{\P}{\text{P}}
\newcommand{\pd}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\sumin}{\sum_i^n}
\newcommand{\Bias}{\text{Bias}}

---

Model too simple: does not fit the data well
– A biased solution


Model too complex: small changes to the data, solution changes a lot
– A high-variance solution


