---
title: "ps4"
author: "whomsoever"
date: "24/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Estimator Bias

This question will guide you through a *simulation study* in `R` to understand the bias of a certain estimator.

```{r}
# This creates a function that calculates sample variance
# but using n in the denominator instead of n-1
# you don't need to change anything here
varn <- function(x) mean((x - mean(x))^2)
```

## Part a.

The code below generates a sample of size n and calculates both the sample variance and the biased version of sample version (the one that has n in the denominator instead of n-1). Modify the code to change n from 10 to some other value between 10 and 50. 

```{r}
# Create a sample of data by rolling a 6-sided die n times
n <- 10000
data <- sample(1:6, n, replace = TRUE)
data
# True variance
35/12
# Unbiased estimate of variance
var(data)
# Biased estimate of variance
varn(data)
```

## Part b.
Repeat the previous experiment many times and find the expected value of each variance estimator.

```{r}
# Unbiased estimator
mean(replicate(10000, var(sample(1:6, 100, replace = TRUE))))
# Biased estimator
mean(replicate(10000, varn(sample(1:6, 1000, replace = TRUE))))
```

## Part c.
After running the previous code, which estimator's average value in the simulation is closer to the true value (roughly 2.9167)? 

\ 

Delete this line and type your answer here.


## Part d.
Now go back and change n to another, larger value, and rerun all of the code. Do you notice anything about the average of the biased estimator?

\ 

Delete this line and type your answer here.

## Part e.
Copy the code from part (b) and paste it below here, then change the `mean` function to be `sd` instead.

```{r}
sd(replicate(10000, var(sample(1:6, 300, replace = TRUE))))
# Biased estimator
sd(replicate(10000, varn(sample(1:6, 300, replace = TRUE))))
```

Delete this sentence and replace with your answer: which estimator has more variability, the biased or unbiased one?




\ 

# Discrimination? 





